---
title: "Gymnast Gold Medal Prediction"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo    = T, 
                      include = T,
                      warning = F,
                      message = F,
                      comment = NA,
                      fig.height = 5, 
                      fig.width  = 5, 
                      fig.align  = 'center')
  library(dplyr)
  library(lubridate)
  library(tidyverse)
  library(tools)
```

## Overview

This project aims to predict the possibility of a gymnast winning a gold medal in different apparatus during finals. We use logistic regression for this prediction.

## Data Exploration
```{r}
# separated qual and final rounds, removed NA score, separated by apparatus
# I believe UE is the same apparatus as UB (uneven bars) UE is not listed in the README 

library('dplyr')
library('lubridate')
library("tidyverse")

x <- read.csv("data/data_2017_2021.csv", as.is = TRUE)
formats = c("%d/ %b/ %Y", "%d/ %B/ %Y", "%a/ %d/ %b/ %Y")
x$Date <- parse_date_time(x$Date, formats)
x <- x %>% 
  distinct() %>% 
  drop_na(Score) 

female_17_21 <- filter(x, Gender == "w")
male_17_21 <- filter(x, Gender == "m")


#rm(list = ls())
apparatus_m <- unique(male_17_21$Apparatus)
apparatus_f <- unique(female_17_21$Apparatus)

# Separating all USA 2017-2021 data by apparatus and round
# Currently there is no male data


# BB women only
f17_21_BB_qual <- filter(female_17_21, Apparatus == "BB" & Round == "qual")
f17_21_BB_final <- filter(female_17_21, Apparatus == "BB" & Round == "final")


# FX women only
f17_21_FX_qual <- filter(female_17_21, Apparatus == "FX" & Round == "qual")
f17_21_FX_final <- filter(female_17_21, Apparatus == "FX" & Round == "final")


# UE women only
f17_21_UE_qual <- filter(female_17_21, Apparatus == "UE" & Round == "qual")
f17_21_UE_final <- filter(female_17_21, Apparatus == "UE" & Round == "final")


# VT both men and women
f17_21_VT_qual <- filter(female_17_21, Apparatus == "VT" & Round == "qual")
f17_21_VT_final <- filter(female_17_21, Apparatus == "VT" & Round == "final")
m17_21_VT_qual <- filter(male_17_21, Apparatus == "VT" & Round == "qual")
m17_21_VT_final <- filter(male_17_21, Apparatus == "VT" & Round == "final")

# VT1 both men and women
f17_21_VT1_qual <- filter(female_17_21, Apparatus == "VT1" & Round == "qual")
f17_21_VT1_final <- filter(female_17_21, Apparatus == "VT1" & Round == "final")
m17_21_VT1_qual <- filter(male_17_21, Apparatus == "VT1" & Round == "qual")
m17_21_VT1_final <- filter(male_17_21, Apparatus == "VT1" & Round == "final")

# VT2 both men and women
f17_21_VT2_qual <- filter(female_17_21, Apparatus == "VT2" & Round == "qual")
f17_21_VT2_final <- filter(female_17_21, Apparatus == "VT2" & Round == "final")
m17_21_VT2_qual <- filter(male_17_21, Apparatus == "VT2" & Round == "qual")
m17_21_VT2_final <- filter(male_17_21, Apparatus == "VT2" & Round == "final")

# SR men only
m17_21_SR_qual <- filter(male_17_21, Apparatus == "SR" & Round == "qual")
m17_21_SR_final <- filter(male_17_21, Apparatus == "SR" & Round == "final")

# PH men only
m17_21_PH_qual <- filter(male_17_21, Apparatus == "PH" & Round == "qual")
m17_21_PH_final <- filter(male_17_21, Apparatus == "PH" & Round == "final")

# PB men only
m17_21_PB_qual <- filter(male_17_21, Apparatus == "PB" & Round == "qual")
m17_21_PB_final <- filter(male_17_21, Apparatus == "PB" & Round == "final")

# HB men only (equiv to FX for women)
m17_21_HB_qual <- filter(male_17_21, Apparatus == "HB" & Round == "qual")
m17_21_HB_final <- filter(male_17_21, Apparatus == "HB" & Round == "final")


y <- read.csv("data/data_2022_2023.csv", as.is = TRUE)
y <- y %>% 
  distinct() %>% 
  drop_na(Score) 

female_22_23 <- filter(y, Gender == "w")
male_22_23 <- filter(y, Gender == "m")

#rm(list = ls())
apparatus_m2 <- unique(male_22_23$Apparatus)
apparatus_f2 <- unique(female_22_23$Apparatus)



# Separating all USA 2022-2023 data by apparatus and round

# BB women only
f22_23_BB_qual <- filter(female_22_23, Apparatus == "BB" & Round == "qual")
f22_23_BB_final <- filter(female_22_23, Apparatus == "BB" & Round == "final")


# FX women only
f22_23_FX_qual <- filter(female_22_23, Apparatus == "FX" & Round == "qual")
f22_23_FX_final <- filter(female_22_23, Apparatus == "FX" & Round == "final")


# UE women only
f22_23_UE_qual <- filter(female_22_23, Apparatus == "UE" & Round == "qual")
f22_23_UE_final <- filter(female_22_23, Apparatus == "UE" & Round == "final")


# VT both men and women
f22_23_VT_qual <- filter(female_22_23, Apparatus == "VT" & Round == "qual")
f22_23_VT_final <- filter(female_22_23, Apparatus == "VT" & Round == "final")
m22_23_VT_qual <- filter(male_22_23, Apparatus == "VT" & Round == "qual")
m22_23_VT_final <- filter(male_22_23, Apparatus == "VT" & Round == "final")

# VT1 both men and women
f22_23_VT1_qual <- filter(female_22_23, Apparatus == "VT1" & Round == "qual")
f22_23_VT1_final <- filter(female_22_23, Apparatus == "VT1" & Round == "final")
m22_23_VT1_qual <- filter(male_22_23, Apparatus == "VT1" & Round == "qual")
m22_23_VT1_final <- filter(male_22_23, Apparatus == "VT1" & Round == "final")

# VT2 both men and women
f22_23_VT2_qual <- filter(female_22_23, Apparatus == "VT2" & Round == "qual")
f22_23_VT2_final <- filter(female_22_23, Apparatus == "VT2" & Round == "final")
m22_23_VT2_qual <- filter(male_22_23, Apparatus == "VT2" & Round == "qual")
m22_23_VT2_final <- filter(male_22_23, Apparatus == "VT2" & Round == "final")

# SR men only
m22_23_SR_qual <- filter(male_22_23, Apparatus == "SR" & Round == "qual")
m22_23_SR_final <- filter(male_22_23, Apparatus == "SR" & Round == "final")

# PH men only
m22_23_PH_qual <- filter(male_22_23, Apparatus == "PH" & Round == "qual")
m22_23_PH_final <- filter(male_22_23, Apparatus == "PH" & Round == "final")

# PB men only
m22_23_PB_qual <- filter(male_22_23, Apparatus == "PB" & Round == "qual")
m22_23_PB_final <- filter(male_22_23, Apparatus == "PB" & Round == "final")

# HB men only (equiv to FX for women)
m22_23_HB_qual <- filter(male_22_23, Apparatus == "HB" & Round == "qual")
m22_23_HB_final <- filter(male_22_23, Apparatus == "HB" & Round == "final")

```


Find the trajectories of D score and E score:
```{r}

## for female_17_21


ggplot(female_17_21, aes(x=LastName)) + 
  geom_point(aes(y = D_Score, color = "D_Score")) + 
  geom_point(aes(y = E_Score, color = "E_Score")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  xlab("Olympian last name") + ylab("Score") + 
  scale_color_manual(name = 'Score labels', breaks = c("D_Score", "E_Score"),
                     values = c("D_Score" = "darkred", "E_Score" = "steelblue"))

```
The graph for the comparison of the D score and E score amongst the gymnasts. We can see the general trends; Biles shows strong lead; Chiles, MC and Carey have greater range 


```{r}


## for female_22_23

ggplot(female_22_23, aes(x=LastName)) + 
  geom_point(aes(y = D_Score, color = "D_Score")) + 
  geom_point(aes(y = E_Score, color = "E_Score")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  xlab("Olympian last name") + ylab("Score") + 
  scale_color_manual(name = 'Score labels', breaks = c("D_Score", "E_Score"),
                     values = c("D_Score" = "darkred", "E_Score" = "steelblue"))

```
Interestingly, we note that the scores are more centralized and have less variance; also interesting to note that Wont has a higher variance compared others; Roberson has a more clusterd datapoints 



```{r}

## for male_22_23
ggplot(male_22_23, aes(x=LastName)) + 
  geom_point(aes(y = D_Score, color = "D_Score")) + 
  geom_point(aes(y = E_Score, color = "E_Score")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  xlab("Olympian last name") + ylab("Score") + 
  scale_color_manual(name = 'Score labels', breaks = c("D_Score", "E_Score"),
                     values = c("D_Score" = "darkred", "E_Score" = "steelblue"))

```
Note that the male_17_21 is missing elements; we briefly talked about this with Brian beforehand as well. Off the bat, men have competed in more competitions than the women, so we have more data for the men's side ; interesting to see how some men have 0 E_score; Malone has good consistency for D score and so does Richard; Whittenburg has less variance for E score but higher variance for D score



```{r}

## Find the distribution of final scores:
ggplot(female_17_21, aes(x=LastName)) + 
  geom_point(aes(y = Score), color = "darkred") 
# Carey's scoes are the most varied; Biles is the most consitently high

ggplot(female_22_23, aes(x=LastName)) + 
  geom_point(aes(y = Score), color = "darkred") 
# Significant improvement in scores; not sure about Matthews, Miller, and 
# Sullivan with the scores of 0
# Miller's scores are very varies; Carey and Jones are consistently well 

ggplot(male_22_23, aes(x=LastName)) + 
  geom_point(aes(y = Score), color = "darkred") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
# Male scores seem more varied; there seems to be higher variance for Hong and 
# Phillips ; some players like 4Nedoroscik, Sun, and Skirkey have little 
# experience and data on them- maybe we can use this to determine who to 
# consider for the team

```

### Variables

The predictor variables used include:
- **Mean Score**: The average score from past competitions.
- **Consistency**: Variance of scores to indicate how consistent a gymnast is.
- **Mean Difficulty**: Average difficulty score chosen by the gymnast.
- **Failure Frequency**: Frequency of penalties in past performances.

Note that all the predictors are based on gymnasts and the apparatus

### Model

We used a logistic regression model for our predictions. 
The summary of the fitted model is here:
```{r}
setwd("/Users/guanrui/Desktop/S&DS625/Gymnastics")

source("clean_data.r")
source("craft_variables.r")
source("fit_model.r")

cleaned_data <- clean_championship_data()
crafted_variables <- craft_variables(cleaned_data)
fit_full_logistic_model(crafted_variables)
```

### Next Steps

1. Incorporate more features, such as coach experience, age of the gymnast, and more.
2. Use other classification models and compare their performance.
3. Build more metrics to evaluate the model.
4. Get more insights according to data visualization
